# Getting Started with R for Data Analysis

The book uses `R` for statistical analyses (<http://www.r-project.org>).
R is a free software environment; you can download it at no charge here: <https://cran.r-project.org>.

## Initial Setup {#sec-initialSetup}

To get started, follow the following steps:

1. Install `R`: <https://cran.r-project.org>
1. Install `RStudio Desktop`: <https://posit.co/download/rstudio-desktop>
1. After installing `RStudio`, open `RStudio` and run the following code in the console to install several key `R` packages:
```{r}
#| eval: false

install.packages(c("petersenlab","tidyverse","psych"))
```

::: {#nte-gettingStarted .callout-note title="If you are in Dr. Petersen's class"}
If you are in Dr. Petersen's class, also perform the following steps:

1. Set up a free account on [GitHub.com](https://github.com).
1. Download GitHub Desktop: <https://desktop.github.com>
:::

## Installing Packages {#sec-installingPackages}

You can install R packages using the following syntax:

```{r}
#| eval: false

install.packages("INSERT_PACKAGE_NAME_HERE")
```
