# Mixed Models {#sec-mixedModels}

## Getting Started {#sec-mixedModelsGettingStarted}

### Load Packages {#sec-mixedModelsLoadPackages}

```{r}

```

## Overview of Mixed Models {#sec-mixedModelsOverview}

## Age-Performance Curve {#sec-agePerformanceCurve}

Fit separately by position:

- random intercepts
- random linear slopes (i.e., random effect of age)
- fixed effects of linear and quadratic age
- fixed effect of experience

## Bayesian Mixed Models {#sec-mixedModelsBayesian}

Model 1 (fit separately by position):

- random intercepts
- random linear slopes (i.e., random effect of age)
- fixed effects of linear and quadratic age
- fixed effect of experience

Model 2 (fit separately by position):

- random intercepts
- random GAM slopes (i.e., age)
- fixed effects of GAM slopes (i.e., age)
- fixed effect of experience

## Conclusion {#sec-mixedModelsConclusion}

::: {.content-visible when-format="html"}

## Session Info {#sec-mixedModelsSessionInfo}

```{r}
sessionInfo()
```

:::
