# Sports and Cognitive Psychology {#sec-sportsCognitivePsychology}

## Getting Started {#sec-sportsCognitivePsychologyGettingStarted}

### Load Packages {#sec-sportsCognitivePsychologyLoadPackages}

```{r}
library("nflreadr")
library("tidyverse")
library("ggtext")
```

### Download Football Data {#sec-sportsCognitivePsychologyDownloadFootballData}

```{r}
#| eval: false
#| include: false

load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_pbp.RData", fsep = ""))
load(file = file.path(path, "/OneDrive - University of Iowa/Teaching/Courses/Fantasy Football/Data/nfl_4thdown.RData", fsep = ""))
```

```{r}
load(file = "./data/nfl_pbp.RData")
load(file = "./data/nfl_4thdown.RData")
```

## Overview {#sec-sportsCognitivePsychologyOverview}

<https://stacker.com/sports/15-ways-analytics-has-changed-sports> (archived at <https://perma.cc/PQ5R-TWFA>)

- Bill James, "Moneyball", Billy Beane, Theo Epstein, and role of Sabermetrics in helping MLB teams with limited budgets (e.g., Oakland Athletics) compete with larger budgets: using statistics to identify player value more accurately, especially for identifying undervalued players
- development of advanced metrics (e.g., value over replacement player, wins above replacement, fielding independent pitching)
- focus on on-base percentage and slugging percentage (and on-base plus slugging percentage) instead of batting average; defensive shifting; fewer attempts to steal bases; fewer bunts; batters taking more pitches; more frequent pitching changes (for pitcher–batter matchups, e.g., righty–lefty); focus on velocity and spin rate
- other sports, including basketball and football are making more use of analytics
- basketball: types of shots (focusing on three pointers and layups); star players resting more games
- football: going for it on fourth down; greater emphasis on the passing game; drafting Running Backs later in the draft (and, overall, valuing Running Backs less); trading down in the draft for more low picks than fewer high picks [because top picks are frequently overvalued; @Massey2013]

## Coaching and Risk Aversion {#sec-coachingRiskAversion}

```{r}
nfl_pbp4thDown <- nfl_pbp %>% 
  filter(down == 4) %>% 
  filter(!(play_type %in% c("no_play","qb_kneel")))

nfl_pbp4thDown$goForIt <- NA
nfl_pbp4thDown$goForIt[which(nfl_pbp4thDown$play_type %in% c("field_goal","punt"))] <- 0
nfl_pbp4thDown$goForIt[which(nfl_pbp4thDown$play_type %in% c("pass","run"))] <- 1

nfl_pbp4thDownPlotData <- nfl_pbp4thDown %>% 
  filter(!is.na(goForIt)) %>% 
  group_by(season) %>% 
  summarise(
    goForItPct = mean(goForIt, na.rm = TRUE),
    n = n(),
    sd = sd(goForIt),
    se = sd / n
  )

ggplot2::ggplot(
  data = nfl_pbp4thDownPlotData,
  ggplot2::aes(
    x = season,
    y = goForItPct)) +
  geom_point() +
  geom_line() +
  geom_ribbon(
    aes(
      y = goForItPct,
      ymin = goForItPct - qnorm(0.975)*se,
      ymax = goForItPct + qnorm(0.975)*se),
    alpha = 0.2) +
  scale_y_continuous(
    limits = c(0, NA)
  ) +
  ggplot2::labs(
    x = "Season",
    y = "Proportion of 4th Down Plays that are Attempts (to Get the First Down)",
    title = "4th Down Attempts (Proportion) by Season",
  ) +
  ggplot2::theme_classic()
```

Adapted from Ben Baldwin: https://www.nfl4th.com/articles/4th-down-research.html (archived at <https://perma.cc/S5D8-3NCU>)

```{r}
# labels on the plot
text_df <- tibble(
  label = c("NFL coaches<br>in <span style='color:#00BFC4'>**2024**</span>", "NFL coaches<br>in <span style='color:#F8766D'>**2014**</span>"),
  x = c(6, 8.2),
  y = c(80, 37),
  angle = c(10, 10),
  color = c("black", "black")
)

nfl_4thdown %>%
  filter(vegas_wp > .2, between(go_boost, -10, 10), season %in% c(2014, 2024)) %>%
  ggplot(
    aes(
      x = go_boost,
      y = go,
      color = as.factor(season))) + 
  ggtext::geom_richtext(
    data = text_df,   
    aes(
      x,
      y,
      label = label,
      angle = angle), 
    color = "black",
    fill = NA,
    label.color = NA,
    size = 5) + 
  geom_vline(xintercept = 0) +
  stat_smooth(
    method = "gam",
    method.args = list(gamma = 1),
    formula = y ~ s(x, bs = "cs", k = 10),
    show.legend = FALSE,
    se = FALSE,
    linewidth = 4) +
  # this is just to get the plot to draw the full 0 to 100 range
  geom_hline(
    yintercept = 100,
    alpha = 0) +
  geom_hline(
    yintercept = 0,
    alpha = 0) +
  ggplot2::theme_classic() +
  labs(
    x = "Gain in win probability by going for it",
    y = "Go-for-it percentage",
    subtitle = "4th down decisions in 2024 versus 2014, win prob. > 20%",
    title = glue::glue("How <span style='color:red'>math</span> is changing football")) +
  theme(
    legend.position = "none",
    plot.title = element_markdown(size = 22, hjust = 0.5),
    plot.subtitle = element_markdown(size = 14, hjust = 0.5),
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold")
  ) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 4),
    limits = c(0, 100),
    expand = c(0,0)) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 10),
    limits = c(-10, 10),
    expand = c(0,0)) +
  annotate(
    "text",
    x = -1.2,
    y = 70,
    label = "Should\nkick",
    color = "black",
    size = 5) +
  annotate(
    "text",
    x = 1.2,
    y = 70,
    label = "Should\ngo for it",
    color = "black",
    size = 5) +
  geom_segment(
    aes(
      x = -.1,
      y = 80,
      xend = -2,
      yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2) +
  geom_segment(
    aes(
      x = .1,
      y = 80,
      xend = 2,
      yend = 80),
    arrow = arrow(length = unit(0.05, "npc")),
    color = "black",
    linewidth = 2)
```

## Conclusion {#sec-sportsCognitivePsychologyConclusion}

::: {.content-visible when-format="html"}

## Session Info {#sec-sportsCognitivePsychologySessionInfo}

```{r}
sessionInfo()
```

:::
